Fail Faster

### 2025-03-24 paper/paper.tex
hyphens

Property-based testing~\cite{claessenQuickCheckLightweightTool2000} is an approach to software testing that centers around
executable specifications of programs called {\em properties}. For example, if a
programmer wants to test an invariant of a binary search tree (BST)
implementation that they are working on, they may write a property like
\begin{center}
  \texttt{prop\_insert\_invariant t x = isBST t ==> isBST (insert x t)}
\end{center}
to check that for any tree \texttt{t} and value \texttt{x}, if \texttt{t} is
already a valid BST then inserting \texttt{x} into it also yields a valid BST.
Once a developer has a property, they test that property by executing
it on hundreds or thousands of random test inputs. These test inputs are usually
produced by a {\em generator} --- a program written in some domain-specific
language (DSL) that allows the developer to express precisely how values should
be sampled.

@@

It might be surprising to readers that this dramatic overhead exists --- shouldn't the compiler perform
this simple optimization? Compilers of effectful and strict functional languages (including JIT
compilers in the case of Scala 3)
do in fact have heuristics to determine if and when to perform this particular kind of
simplification.\footnote{As we discuss in Section~\ref{section:other-langs}, purity
means that Haskell is a slightly different story. GHC can and often does
transformations of this form.} However even in cases as simple as
Figure~\ref{fig:simplify}, the indirection of \texttt{return}
and \texttt{bind} causes these heuristics to not fire.  Moreover, the story is
even worse for recursive generators, as the heuristics are necessarily even more
conservative for optimizing recursive functions.

@@

Last, many PBT libraries --- including both \bq and ScalaCheck --- implement their \texttt{weighted\_union} combinators
in ways that are asymptotically more efficient but slower in common cases than a more naive algorithm.
Weighted union uses the Fitness Proportionate Selection algorithm~\cite{geneticalgorithm}, which (1) samples a number $r$
between $0$ and the sum of the weights, and then (2) finds the first generator for which the cumulative
weights in the list before it exceeds $r$. This second part can be accomplished in $O(\log_2 n)$ time by
a binary search. However, since the lists are short in practice, a linear scan is almost always faster.
Moreover, both \bq{} and ScalaCheck allocate auxiliary data structures (an array in \bq{} and a BST in ScalaCheck) to
perform this search, which incurs further run time overhead.

@@

The randomness library is the hottest part of the hot path.
Indeed, even basic generators --- like ones generating a single \texttt{int} or \texttt{float} uniformly within a range --- can sample \emph{unboundedly many}
random numbers, since they usually use versions of rejection sampling~\cite{vn51} to find a value within the range.
Moreover, generator combinators like \texttt{list} usually make $O(n)$ calls to the \texttt{int} or \texttt{float} generators.
Because of this, the speed of a single sample matters a great deal. Unfortunately,
while existing PBT libraries by and large make sensible choices for their randomness libraries,
they are not chosen with performance in mind, leading to worse bugfinding power than what is possible.

@@

The power of staging for optimizing away abstraction overheads comes from
defining functions that accept and return \texttt{code} values.  A function
\texttt{f : 'a code -> 'b code} takes a program computing a
run-time \texttt{'a} and transforms it into a program computing a run-time
\texttt{'b}. In particular, because \texttt{f} itself runs at compile time, the
fact that the programmer called \texttt{f} does not matter at run time---the abstraction
that \texttt{f} defines has been eliminated.
A code-transforming function \texttt{'a code -> 'b code} can also be converted \emph{code for a function} --- a value of type
\texttt{('a -> 'b) code}---with the following program:
\begin{lstlisting}
let eta (f : 'a code -> 'b code) : ('a -> 'b) code = .<fun x -> .~(f .<x>.)>.
\end{lstlisting}

@@

% carefully break up so that you push what you can into the compile time stage.
% Figuring out which parts of the generator are known statically ()
% Recall \jwc{did we talk about this} that staging a DSL involves changing the
% combinators to run at compile time by carefully annotating their types with \texttt{code}s.
% Deciding which types \texttt{t} can
% instead be \texttt{t code} --- in other words,
% figuring out which parts of the DSL can be determined statically (and can be
% part of the compile-time stage), and which parts are only known dynamically (and
% hence must be \texttt{code}) --- is an art known as ``binding-time analysis''

@@

The basic functionality of the staged monadic generator DSL can be
found in Figure~\ref{fig:gen-staged-basic}.
The constant generator \texttt{return} runs at compile time. Given \texttt{cx : 'a code}, the code for an \texttt{'a}, it
returns the generator that ignores its \texttt{size} and
\texttt{random} arguments and simply runs\bcp{returns the code for?} \texttt{cx}. Similarly,
\texttt{bind g k} sequences generators by passing the result of running the generator \texttt{g} to the continuation \texttt{k}. However, instead of getting access
to the particular value generated by \texttt{g}, the continuation
\texttt{k} only gets access to \texttt{code} for the value sampled from \texttt{g}:
we know that at run time the code \texttt{g} generates will produce \emph{some} \texttt{'a}, but at compile time we cannot inspect the value.
Operationally, bind takes code for the size and seed and returns code that (1) let-binds a variable \texttt{a} to spliced-in code of \texttt{g} and then
(2) runs the spliced-in continuation \texttt{k}.
Both function applications \texttt{g size random}
and \texttt{k .<a>. size random} run at compile time.
\texttt{Gen.int} is the generator that samples an int from the randomness library.
Given any size and random seed, it returns a code block that calls \texttt{SR.int} with that random seed. Because the lower and upper bounds
might not be known at compile time --- they may themselves be the results of calling \texttt{Gen.int} ---
the arguments \texttt{lo} and \texttt{hi} are of type \texttt{int code} and get spliced into the code block as arguments to \texttt{SR.int}.
Lastly, \texttt{to\_bq} turns a staged generator into code for a normal \bq\
generator. This is just a 2-argument version of ``The Trick''
from Section~\ref{subsection:msp}.
\ifaftersubmission
\hg{TODO: Be really careful with formatting of the above paragraph. It's very
easy for some inline code to get split weirdly over a line break or just be
difficult to parse in general, and that could throw the reader off when the
content is already pretty low-level}
\fi

@@

In Section~\ref{section:motiv}, we noted that generator combinators like \texttt{weighted\_union}
allocate lists in the hot path of the generator. Even though these lists are usually small ---
at most a few dozen elements --- each allocation takes us closer to the next garbage collection, which is bad for performance.

@@

Figure~\ref{fig:staged-weighted-union} shows the code for such a staged weighted union.
Crucially, it takes a \emph{compile-time} list \texttt{weighted\_gens} of
generators and weights. The weights themselves might only be known at run time --- it is
common to use the current size parameter as a weight, for instance ---
so they are \texttt{code}s.
Instead of building a data structure representing a histogram of the distribution described by the weights at run time and then traversing it,
the compile-time \texttt{weighted\_union} combinator generates a
tree of \texttt{if}s, specialized to the list of weights known at compile time, that picks out the selected generator.

@@

% \jwc{Describe split --- this section is approximately experts-only, but you can just point at the Andras paper and the things it cites.}

@@

In the years since the original QuickCheck paper~\cite{quickcheck00}, PBT has had remarkable success
in languages outside of the Haskell world from which it came~\cite{bq,sc,hypothesis}.
For this reason, readers who are users and developers of PBT libraries in \emph{other} languages ---
including non-strict or procedural ones --- might be curious about how the
techniques brought to bear on OCaml's \bq{} and Scala's ScalaCheck in this paper
might be imported to their favorite language.

@@

  \textbf{Racket} is the next target that we intend to test on.
  The entire Racket philosophy is intertwined with
  using macros to build small eDSLs like the ones we use to write generators, and so it seems a natural fit.
  However, the racket macros literature, while extensive, does not usually concern itself with staging for performance purposes.
  For this reason, we --- the authorship team of this paper, devoid of much Racket expertise --- decided to not use Racket as a test case in this paper.

@@

  \textbf{Rust} shares some similarities with functional languages, and
  indeed, it is host to a number of property-based testing libraries~\cite{Bolero,ProptestBook}. However Rust does not
  structurally encourage monadic eDSLs --- it does not have a special monadic syntax --- and so
  PBT libraries in Rust ask programmers to define generators directly as \texttt{seed -> value} functions.
  For this reason, staging does not seem directly applicable to any of the PBT libraries in Rust.

>>>>

Property-based testing~\cite{claessenQuickCheckLightweightTool2000} is an approach to software testing that centers around
executable specifications of programs called {\em properties}. For example, if a
programmer wants to test an invariant of a binary search tree (BST)
implementation that they are working on, they may write a property like
\begin{center}
  \texttt{prop\_insert\_invariant t x = isBST t ==> isBST (insert x t)}
\end{center}
to check that for any tree \texttt{t} and value \texttt{x}, if \texttt{t} is
already a valid BST then inserting \texttt{x} into it also yields a valid BST.
Once a developer has a property, they test that property by executing
it on hundreds or thousands of random test inputs. These test inputs are usually
produced by a {\em generator}---a program written in some domain-specific
language (DSL) that allows the developer to express precisely how values should
be sampled.

@@

It might be surprising to readers that this dramatic overhead exists---shouldn't the compiler perform
this simple optimization? Compilers of effectful and strict functional languages (including JIT
compilers in the case of Scala 3)
do in fact have heuristics to determine if and when to perform this particular kind of
simplification.\footnote{As we discuss in Section~\ref{section:other-langs}, purity
means that Haskell is a slightly different story. GHC can and often does
transformations of this form.} However even in cases as simple as
Figure~\ref{fig:simplify}, the indirection of \texttt{return}
and \texttt{bind} causes these heuristics to not fire.  Moreover, the story is
even worse for recursive generators, as the heuristics are necessarily even more
conservative for optimizing recursive functions.

@@

Last, many PBT libraries---including both \bq and ScalaCheck---implement their \texttt{weighted\_union} combinators
in ways that are asymptotically more efficient but slower in common cases than a more naive algorithm.
Weighted union uses the Fitness Proportionate Selection algorithm~\cite{geneticalgorithm}, which (1) samples a number $r$
between $0$ and the sum of the weights, and then (2) finds the first generator for which the cumulative
weights in the list before it exceeds $r$. This second part can be accomplished in $O(\log_2 n)$ time by
a binary search. However, since the lists are short in practice, a linear scan is almost always faster.
Moreover, both \bq{} and ScalaCheck allocate auxiliary data structures (an array in \bq{} and a BST in ScalaCheck) to
perform this search, which incurs further run time overhead.

@@

The randomness library is the hottest part of the hot path.
Indeed, even basic generators---like ones generating a single \texttt{int} or \texttt{float} uniformly within a range---can sample \emph{unboundedly many}
random numbers, since they usually use versions of rejection sampling~\cite{vn51} to find a value within the range.
Moreover, generator combinators like \texttt{list} usually make $O(n)$ calls to the \texttt{int} or \texttt{float} generators.
Because of this, the speed of a single sample matters a great deal. Unfortunately,
while existing PBT libraries by and large make sensible choices for their randomness libraries,
they are not chosen with performance in mind, leading to worse bugfinding power than what is possible.

@@

The power of staging for optimizing away abstraction overheads comes from
defining functions that accept and return \texttt{code} values.  A function
\texttt{f : 'a code -> 'b code} takes a program computing a
run-time \texttt{'a} and transforms it into a program computing a run-time
\texttt{'b}. In particular, because \texttt{f} itself runs at compile time, the
fact that the programmer called \texttt{f} does not matter at run time---the abstraction
that \texttt{f} defines has been eliminated.
A code-transforming function \texttt{'a code -> 'b code} can also be converted \emph{code for a function}---a value of type
\texttt{('a -> 'b) code}---with the following program:
\begin{lstlisting}
let eta (f : 'a code -> 'b code) : ('a -> 'b) code = .<fun x -> .~(f .<x>.)>.
\end{lstlisting}

@@

% carefully break up so that you push what you can into the compile time stage.
% Figuring out which parts of the generator are known statically ()
% Recall \jwc{did we talk about this} that staging a DSL involves changing the
% combinators to run at compile time by carefully annotating their types with \texttt{code}s.
% Deciding which types \texttt{t} can
% instead be \texttt{t code}---in other words,
% figuring out which parts of the DSL can be determined statically (and can be
% part of the compile-time stage), and which parts are only known dynamically (and
% hence must be \texttt{code})---is an art known as ``binding-time analysis''

@@

The basic functionality of the staged monadic generator DSL can be
found in Figure~\ref{fig:gen-staged-basic}.
The constant generator \texttt{return} runs at compile time. Given \texttt{cx : 'a code}, the code for an \texttt{'a}, it
returns the generator that ignores its \texttt{size} and
\texttt{random} arguments and simply runs\bcp{returns the code for?} \texttt{cx}. Similarly,
\texttt{bind g k} sequences generators by passing the result of running the generator \texttt{g} to the continuation \texttt{k}. However, instead of getting access
to the particular value generated by \texttt{g}, the continuation
\texttt{k} only gets access to \texttt{code} for the value sampled from \texttt{g}:
we know that at run time the code \texttt{g} generates will produce \emph{some} \texttt{'a}, but at compile time we cannot inspect the value.
Operationally, bind takes code for the size and seed and returns code that (1) let-binds a variable \texttt{a} to spliced-in code of \texttt{g} and then
(2) runs the spliced-in continuation \texttt{k}.
Both function applications \texttt{g size random}
and \texttt{k .<a>. size random} run at compile time.
\texttt{Gen.int} is the generator that samples an int from the randomness library.
Given any size and random seed, it returns a code block that calls \texttt{SR.int} with that random seed. Because the lower and upper bounds
might not be known at compile time---they may themselves be the results of calling \texttt{Gen.int}---the arguments 
\texttt{lo} and \texttt{hi} are of type \texttt{int code} and get spliced into the code block as arguments to \texttt{SR.int}.
Lastly, \texttt{to\_bq} turns a staged generator into code for a normal \bq\
generator. This is just a 2-argument version of ``The Trick''
from Section~\ref{subsection:msp}.
\ifaftersubmission
\hg{TODO: Be really careful with formatting of the above paragraph. It's very
easy for some inline code to get split weirdly over a line break or just be
difficult to parse in general, and that could throw the reader off when the
content is already pretty low-level}
\fi

@@

In Section~\ref{section:motiv}, we noted that generator combinators like \texttt{weighted\_union}
allocate lists in the hot path of the generator. Even though these lists are usually small---at most a few dozen elements---each allocation takes us closer to the next garbage collection, which is bad for performance.

@@

Figure~\ref{fig:staged-weighted-union} shows the code for such a staged weighted union.
Crucially, it takes a \emph{compile-time} list \texttt{weighted\_gens} of
generators and weights. The weights themselves might only be known at run time---it is
common to use the current size parameter as a weight, for instance---so they are \texttt{code}s.
Instead of building a data structure representing a histogram of the distribution described by the weights at run time and then traversing it,
the compile-time \texttt{weighted\_union} combinator generates a
tree of \texttt{if}s, specialized to the list of weights known at compile time, that picks out the selected generator.

@@

% \jwc{Describe split---this section is approximately experts-only, but you can just point at the Andras paper and the things it cites.}

@@

In the years since the original QuickCheck paper~\cite{quickcheck00}, PBT has had remarkable success
in languages outside of the Haskell world from which it came~\cite{bq,sc,hypothesis}.
For this reason, readers who are users and developers of PBT libraries in \emph{other} languages---including non-strict or procedural ones---might be curious about how the
techniques brought to bear on OCaml's \bq{} and Scala's ScalaCheck in this paper
might be imported to their favorite language.

@@

  \textbf{Racket} is the next target that we intend to test on.
  The entire Racket philosophy is intertwined with
  using macros to build small eDSLs like the ones we use to write generators, and so it seems a natural fit.
  However, the racket macros literature, while extensive, does not usually concern itself with staging for performance purposes.
  For this reason, we---the authorship team of this paper, devoid of much Racket expertise---decided to not use Racket as a test case in this paper.

@@

  \textbf{Rust} shares some similarities with functional languages, and
  indeed, it is host to a number of property-based testing libraries~\cite{Bolero,ProptestBook}. However Rust does not
  structurally encourage monadic eDSLs---it does not have a special monadic syntax---and so
  PBT libraries in Rust ask programmers to define generators directly as \texttt{seed -> value} functions.
  For this reason, staging does not seem directly applicable to any of the PBT libraries in Rust.
